---
title: "Introduction to single-cell RNA-seq analysis"
subtitle: "Cluster Marker Genes"
output:
  html_document:
    toc: yes
    toc_float: true
    toc_depth: 3
    number_sections: true
    pandoc_args: ["--number-offset", "8"] # TOC will start at 9
    code_folding: show
    css: ../css/boxes.css
---

# Introduction

In order to aid the interpretation of the clustering results that we covered in
the [previous section](08_Clustering.html), it is helpful to identify genes that
contribute to the separation of cells into those clusters.

The main approach to achieve this, is to identify genes that are differently
expressed between clusters. These may be, for example, exclusively expressed in
a single cluster or perhaps differentiate between a few different clusters.
There are different methods to identify expression differences between clusters:
using mean expression level, or the ranking of the gene by its expression, or
the proportions of cells that express the gene.

Our main objective in this section is to cover some of the methods that can be used to achieve this goal, and obtain a summary table of results.
As always, the [OSCA chapter on marker detection](https://bioconductor.org/books/release/OSCA.basic/marker-detection.html)
contains additional detail and advice.

## Load Data

```{r libraries, message=FALSE, warning=FALSE}
library(Seurat)
library(sctransform)
library(glmGamPoi)
library(tidyverse)
library(patchwork)
```

First we will load in a version of our data that has been processed up until this point in the course. QCed, normalised, dimensionality reduced and clustered. 

```{r data}
seurat_object <- readRDS("../RObjects/Clustered.500.rds")
```
The cluster information is stored in our objects `meta.data` slot and is called "leiden_cluster". 

To remind ourselves, we can visualise the clusters on a UMAP:

```{r umap}
DimPlot(seurat_object,
        reduction = "umap",
        group.by = "leiden_cluster",
        label = TRUE,
        label.size = 5) + NoLegend()
```

Our objective is to identify genes that distinguish these clusters from one
another - "cluster marker genes". Intuitively we hope that the clusters relate
to specific cell populations, and therefore we are trying to find genes that
will allow us to identify the cell types for each cluster.

For example genes such as the "CST3" gene, which is a known monocyte marker:

```{r cst3}
# visualise known monocyte marker gene CST3
FeaturePlot(seurat_object,
            reduction = "umap",
            features = "CST3")
```

## Identifying cluster marker genes

Although we have defined our clusters based on the batch-corrected expression
values, these should **not** be used for for gene-based analyses like marker
gene detection. Instead, we should **use the uncorrected (normalised) expression
values for differential expression between clusters**. This is because data
integration algorithms bring cells together based on their overall gene
expression, but for each gene individually the data transformation may introduce
artificial agreement between batches, which is not ideal for gene-level
differential analysis. Furthermore, the severity of these biases is dependent on
the parameters used for data integration.

Valid assays to use in gene based differential analysis tests are the normalised
counts obtained from the SCTransform methodand stored int he "SCT" assay. 

Seurat has two functions we can use for this, `FindMarkers()` and `FindAllMarkers()`. The former allows us to compare two clusters at a time, in a pairwise manner. The latter automates this for all clusters but can also be used to test groups of clusters against other clusters or against all cells.

Prior to performing differential expression, we first run PrepSCTFindMarkers. 

"This allows us to use the ‘corrected counts’ that are stored in the data slot of the the SCT assay. Corrected counts are obtained by setting the sequencing depth for all the cells to a fixed value and reversing the learned regularized negative-binomial regression model. PrepSCTFindMarkers() ensures that the fixed value is set properly."
  -- Seurat Vignette

```{r prepmarkers}
seurat_object <- PrepSCTFindMarkers(seurat_object, assay = "SCT")
```


### Finding markers of a single cluster

To find markers of a single cluster, we can use `FindMarkers()`. For example, to find genes that are differentially expressed in cluster 2 compared to all other clusters:

```{r findmarkers2}
# takes about 6 mins to run without presto (teabreak?)
markers_cluster2 <- FindMarkers(seurat_object,
                               ident.1 = 2,
                               ident.2 = NULL,
                               assay = "SCT",
                               group.by = "leiden_cluster")
markers_cluster2 %>% head()
```

The output is a data frame with the following columns:
- `p_val`: p-value of the test
- `avg_log2FC`: average log2 fold change between the two groups
- `pct.1`: percentage of cells in cluster 2 that express the gene
- `pct.2`: percentage of cells in the other clusters that express the gene
- `p_val_adj`: adjusted p-value for multiple testing

Using pvalues is inappropriate for marker gene detection. This is because the differential expression is performed on the same data used to obtain the clusters and would therefore be "data dredging". If we have used gene expression to seperate the cells into clusters we would expect thre to be a difference in expression between clusters. Admitedly, if simply used to identify possible marker genes by ranking this is largely harmless but it becuse problematic when the pvalues are used to claim statistically significant seperation between clusters. There is also an issue of pseudoreplication where in this type of differential application cells from the same cluster are treated as replicates when infact cells from the same sample are not independent. There is a more detailed discussion of these issues here. https://bioconductor.org/books/3.14/OSCA.advanced/marker-detection-redux.html

Instead, we should use the log fold change and percentage of cells expressing the gene to identify marker genes. For example, we could filter for genes that are expressed in at least 70% of cells in cluster 2 and have a log fold change greater than 1:

```{r filtermarkers2}
markers_cluster2_filtered <- markers_cluster2 %>%
  filter(pct.1 > 0.70, abs(avg_log2FC) > 1)
markers_cluster2_filtered %>% head()
```

We can plot some of these marker genes on the UMAP.

```{r markerPlots}
Fplot <- FeaturePlot(seurat_object, features = "SKAP1", reduction = "umap")
Vplot <- VlnPlot(seurat_object, features = "SKAP1", 
                 group.by = "leiden_cluster",
                 pt.size = 0) + NoLegend()
Fplot + Vplot + plot_layout(ncol = 2)
# SKAP1 is a TCR gene
# Also appears in cluster 9 which if we look back at UMAP by cluster is next to it.
```

### Finding markers of all clusters

To do the same process but for all clusters we use `FindAllMarkers()`. 

```{r findallmarkers}

markers_all <- FindAllMarkers(seurat_object,
                              assay = "SCT",
                              group.by = "leiden_cluster")
markers_all %>% head()
```

The output is a data frame with the following columns:
- `gene`: gene name
- `cluster`: cluster number
- `p_val`: p-value of the test
- `avg_log2FC`: average log2 fold change between the cluster and all other clusters
- `pct.1`: percentage of cells in the cluster that express the gene
- `pct.2`: percentage of cells in the other clusters that express the gene
- `p_val_adj`: adjusted p-value for multiple testing

We can filter the dataframe to get the same information as we did before. To get the marker genes for cluster 2:

```{r filterallmarkers2}
markers_all_cluster2 <- markers_all %>%
  filter(cluster == 2, pct.1 > 0.70, abs(avg_log2FC) > 1)
markers_all_cluster2 %>% head()
```

We can see we have the same 29 genes.

## Exercise 1

:::exercise

We saw earlier that known monocyte marker CST3 was expressed in cluster 13. Use the FindMarkers() function or the FindAllMarkers() function to investigate what other potentially good marker genes would be for that cluster.

<details><summary>Answer</summary>

```{r exercise1, purl=FALSE}
# using FindMarkers()
markers_cluster13 <- FindMarkers(seurat_object,
                               ident.1 = 13,
                               ident.2 = NULL,
                               assay = "SCT",
                               group.by = "leiden_cluster")
markers_cluster13_filtered <- markers_cluster13 %>%
  filter(pct.1 > 0.70, abs(avg_log2FC) > 1)
         
markers_cluster13_filtered %>% head()

```

</details>

## Heatmap of marker genes

We have already seen how we can use the `VlnPlot()` function to visualise
the distribution of expression in our data between clusters. We have also seen
how to use `FeaturePlot()` to visualise a gene's expression on the projected
reduced dimensionality space.

Another useful type of visualisation is to use **heatmaps** to show the
expression of these genes of interest. 
We will demonstrate this using the top marker genes for cluster 13. Lets take the top 20 genes from our filtered markers.

```{r heatmap}
# top 20 marker genes for cluster 13
top_markers_cluster13 <- markers_cluster13_filtered %>%
  top_n(n = 20, wt = abs(avg_log2FC)) %>%
  rownames()

# heatmap of expression for these genes

DoHeatmap(seurat_object, 
          features = top_markers_cluster13, 
          group.by = "leiden_cluster") + scale_fill_gradientn(colors = c("navy", "white", "firebrick")) # seurat default colours are ugly




```

## Cell Type Labelling

One of the main tasks we often want to perform is annotating our cells as known types of cells occurring in our sampled tissue. 
This requires prior knowledge of cell transcriptomic states and therefore becomes easier if there are well-curated resources available. 
However, for less well-studied tissues these may not be available and so cell type annotation may rely on "manual annotation" using a small set of genes with known cell-specific expression (e.g. from microscopy data, qPCR on cell-sorted samples, etc.). 

In this section we will do a very simple manual labelling of our clusters, based on known genes expressed in different blood cell types. 
However, there are more systematic methods for cell type annotation, in particular when prior information is available for those cells: 

- The [`SingleR`](https://bioconductor.org/packages/3.16/bioc/vignettes/SingleR/inst/doc/SingleR.html) package uses previously labelled bulk or single-cell datasets to annotate a new dataset.
- The [`AUCcell`](https://bioconductor.org/packages/3.16/bioc/vignettes/AUCell/inst/doc/AUCell.html) package classifies cells into types based on user-provided lists of "signature" genes for each type. These lists can be generated from literature, or also be based on prior RNA-seq datasets. 
- Another strategy is to perform a standard gene set enrichment analysis on the top marker genes for each cluster. 

Any of these approaches is suitable, and they can often be done in conjunction. 
To learn more about these approaches, read the chapter on [cell type annotation in the OSCA book](http://bioconductor.org/books/3.16/OSCA.basic/cell-type-annotation.html).


## Manual annotation

A lot is known about immune cell markers, in particular as many surface markers have been identified as useful for [immunophenotyping](https://en.m.wikipedia.org/wiki/Cluster_of_differentiation#Immunophenotyping). 

To help us in our annotation, we start by retrieving the top-ranked genes from each cluster into a list. 
We do this by looping through the list using the `lapply()` function and in each case picking the genes with rank < 10 for Cohen's D statistic:

```{r all_top_markers}
# loop through all marker genes and extract top-ranked gene names for each cluster
top_markers_all <- lapply(unique(markers_all$cluster), function(x) {
  markers_all %>%
    filter(cluster == x) %>%
    filter(pct.1 > 0.70, abs(avg_log2FC) > 1) %>%
    top_n(n = 15, wt = abs(avg_log2FC)) %>%
    rownames()
})

# examining this list reveals several known markers of immune cells
top_markers_all
```

If we took some time to examine this list (which would benefit from knowing about the underlying biology of the immune system), we can start to see some genes known to differentiate different types of cells: 

- _HBA1_ and _HBA2_ → expressed in red blood cells (erythrocytes)
- _CST3_ → specific to monocytes
- _CD3E_ and _CD3D_ → specific to T cells
- _NKG7_ → specific to natural killer (NK) T cells
- _CD79A_ and _CD24_ → specific to B cells
- _MS4A1_ (CD20) → a clinically-important antigen used as a target to treat B cell-related diseases, such as leukemia. Cells with this antigen are referred to as CD20+ B cells.

Let's visualise these markers' expression in our clusters:

```{r all_marker_violins}
# cell type specific genes
known_genes <- c(
  "HBA1", # erythrocytes
  "CST3", # monocytes
  "CD3E", # T cells
  "NKG7", # NK T cells
  "CD79A",  # B cells
  "MS4A1" # CD20 B cells
  )

# violin plot
VlnPlot(seurat_object,
        features = known_genes,
        group.by = "leiden_cluster",
        pt.size = 0) + NoLegend() + scale_x_discrete(guide =
  guide_axis(angle = 45))

```

Based on the expression patterns for these cells, we classify these clusters as follows: 

| Cluster | Labelled Cell Type                | Evidence      |
| :------ | :-------------------------------- | :------------ |
| 1       | B cells                           | CD79A         |
| 2       | T cells                           | CD79A         |
| 3       | B cells                           | CD79A         |
| 4       | B cells                           | CD79A         |
| 5       | T cells                           | CD3D          |
| 6       | B cells                           | CD79A         |
| 7       | B cells                           | CD79A         |
| 8       | B cells                           | CD79A         |
| 9       | CD20 B cells                      | CD79A + MS4A1 |
| 10      | T cells                           | CD3D          |
| 11      | NK T cells                        | CD3D + NKG7   |
| 12      | B cells                           | CD79A         |
| 13      | Monocytes                         | CD79A         |
| 14      | Erythrocytes                      | HBA1          |
| 15      | Erythrocytes                      | HBA1          |
| 16      | Unknown                           |               |

Note: cluster 16 has not been identified but it's top markers seem to be tumour or cancer related genes, so it may be a cluster of tumour cells.

Now that we have a more meaningful annotation for our clusters, let's add this to our `Seurat` object.
We will also add the original cluster ID in parenthesis to remind ourselves that this annotation was done based on the clusters. 

When you have come to a decision about which clustering to use it is convenient to set `Idents` to the column name of the clustering you have chosen. This means downstream code does not need to be changed should you later decide to switch to a different clustering, you’d simply need to change the contents of the `Idents` slot. This also makes the code easily re-usable for different analyses.

```{r relabel}

temp_metadata <- seurat_object@meta.data %>%
  mutate(Idents = case_when(
    leiden_cluster == 1 ~ "B (c1)",
    leiden_cluster == 2 ~ "T (c2)",
    leiden_cluster == 3 ~ "B (c3)",
    leiden_cluster == 4 ~ "B (c4)",
    leiden_cluster == 5 ~ "T (c5)",
    leiden_cluster == 6 ~ "B (c6)",
    leiden_cluster == 7 ~ "B (c7)",
    leiden_cluster == 8 ~ "B (c8)",
    leiden_cluster == 9 ~ "CD20+ B (c9)",
    leiden_cluster == 10 ~ "T (c10)",
    leiden_cluster == 11 ~ "NK T (c11)",
    leiden_cluster == 12 ~ "B (c12)",
    leiden_cluster == 13 ~ "Monocytes (c13)",
    leiden_cluster == 14 ~ "Erythrocytes (c14)",
    leiden_cluster == 15 ~ "Erythrocytes (c15)",
    leiden_cluster == 16 ~ "Unknown (c16)"
  ))

seurat_object <- AddMetaData(seurat_object, metadata = temp_metadata$Idents, col.name = "Idents")

Idents(seurat_object) <- "Idents"

```

Now, when we label our UMAP, we can see the new names, which are more intutitive to interpret:

```{r umap_relabel}
# visualise UMAP with new labels
DimPlot(seurat_object,
        reduction = "umap",
        group.by = "Idents",
        label = TRUE,
        label.size = 5) + NoLegend()
```

## (Bonus) Further exploration

### Erythrocytes

Looking at the expression of _HBA1_ and _HBA2_: 

```{r, purl=FALSE}

FeaturePlot(seurat_object,
            features = c("HBA1", "HBA2"),
            reduction = "umap") + scale_x_discrete(guide = guide_axis(angle = 45))
```

You may notice that there is a lot of background expression across several cell clusters. 
_HBA1_ and _HBA2_ are common components of the **"soup" or ambient RNA** in scRNA-seq experiments involving blood cells. 
Hemoglobin chains, such as _HBA1_ and _HBA2_, are very highly expressed in red blood cells and burst red blood cells will contribute their RNA to the medium from which the cells are loaded into the 10X Chromium machine. 
Some of this medium, or "soup", containing ambient RNA is captured with each cell. 
There are methods available to correct for soup in your analysis such as [SoupX](https://academic.oup.com/gigascience/article/9/12/giaa151/6049831).

```{r sessioninfo}
sessionInfo()
```


